---
title: "Dissertation"
author: "Dhannyjazz"
date: "10/02/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, message = F, eval = F)
```

##### About the Dataset

GSE35977

##### Install BiocManager and required packages

```{r}
#if (!require("BiocManager", quietly = TRUE))
#  install.packages("BiocManager")

#BiocManager::install("GEOquery")

#BiocManager::install("limma")

#BiocManager::install("dplyr")
```

##### Loading the Dataset

```{r cars}
# Version info: R 3.2.3, Biobase 2.30.0, GEOquery 2.40.0, limma 3.26.8
################################################################
#   Differential expression analysis with limma
library(GEOquery)
library(limma)
library(umap)

# load series and platform data from GEO
gset <- getGEO("GSE35977", GSEMatrix =TRUE, AnnotGPL=TRUE)

if (length(gset) > 1) idx <- grep("GPL6244", attr(gset, "names")) else idx <- 1
gset <- gset[[idx]]

```
```{r}
### make proper column names to match toptable
fvarLabels(gset) <- make.names(fvarLabels(gset))
```

```{r} 
#GET PHENODATA
pheno_dataM <- pData(gset)

head(pheno_dataM)
```

```{r}

```

### group membership for all samples
```{r}
gsms <- paste0("1X0XX0XX11XX0101XXX1X10XX1000XX0011XX00001011001X1",
        "0X100111110X01X0X00X1X011X101X101X0X010110010X10XX",
        "X01XX1000XXXXX10111X000XXXXXX0X0011X0X01XX01X0XXXX",
        "XX1X0111X11XXXXXXX")
sml <- strsplit(gsms, split="")[[1]]
```
FILTER OUT OTHER GROUPS 
#(Unused samples marked as X)
```{r}
sel <- which(sml != "X")
sml <- sml[sel]
gset <- gset[ ,sel]
```

# log2 transformation
```{r}
ex <- exprs(gset)
qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0), na.rm=T))
LogC <- (qx[5] > 100) ||
          (qx[6]-qx[1] > 50 && qx[2] > 0)
if (LogC) { ex[which(ex <= 0)] <- NaN
  exprs(gset) <- log2(ex) }

```

# assign samples to groups and set up design matrix
```{r}
gs <- factor(sml)
groups <- make.names(c("Schz","Control"))
levels(gs) <- groups
gset$group <- gs
design <- model.matrix(~group + 0, gset)
colnames(design) <- levels(gs)

gset <- gset[complete.cases(exprs(gset)), ] # skip missing values

fit <- lmFit(gset, design)  # fit linear model

```
# set up contrasts of interest and recalculate model coefficients
```{r}
cts <- paste(groups[1], groups[2], sep="-")
cont.matrix <- makeContrasts(contrasts=cts, levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)

```

###### compute statistics and table of top significant genes

```{r pressure}
# compute statistics and table of top significant genes
fit2 <- eBayes(fit2, 0.01)
tT <- topTable(fit2, adjust="fdr", sort.by="B", number=Inf)

```


```{r}
tT2 <- topTable(fit2, adjust="fdr", sort.by="B", number=Inf)
hist(tT2$adj.P.Val, col = "grey", border = "white", xlab = "P-adj",
  ylab = "Number of genes", main = "P-adj value distribution")

# summarize test results as "up", "down" or "not expressed"
dT <- decideTests(fit2, adjust.method="fdr", p.value=0.5, lfc=0)


```
```{r}
# Venn diagram of results
vennDiagram(dT, circle.col=palette())


```

```{r}
# create Q-Q plot for t-statistic
t.good <- which(!is.na(fit2$F)) # filter out bad probes
qqt(fit2$t[t.good], fit2$df.total[t.good], main="Moderated t statistic")


```

```{r}
# volcano plot (log P-value vs log fold change)
colnames(fit2) # list contrast names
ct <- 1        # choose contrast of interest
volcanoplot(fit2, coef=ct, main=colnames(fit2)[ct], pch=20,
  highlight=length(which(dT[,ct]!=0)), names=rep('+', nrow(fit2)))

```

```{r}
# MD plot (log fold change vs mean log expression)
# highlight statistically significant (p-adj < 0.05) probes
plotMD(fit2, column=ct, status=dT[,ct], legend=F, pch=20, cex=1)
abline(h=0)

```

```{r}
# General expression data analysis
ex <- exprs(gset)

# box-and-whisker plot
dev.new(width=3+ncol(gset)/6, height=5)
ord <- order(gs)  # order samples by group
palette(c("#1B9E77", "#7570B3", "#E7298A", "#E6AB02", "#D95F02",
          "#66A61E", "#A6761D", "#B32424", "#B324B3", "#666666"))
par(mar=c(7,4,2,1))
title <- paste ("GSE35977", "/", annotation(gset), sep ="")
boxplot(ex[,ord], boxwex=0.6, notch=T, main=title, outline=FALSE, las=2, col=gs[ord])
legend("topleft", groups, fill=palette(), bty="n")
dev.off()

# expression value distribution
par(mar=c(4,4,2,1))
title <- paste ("GSE35977", "/", annotation(gset), " value distribution", sep ="")
plotDensities(ex, group=gs, main=title, legend ="topright")

```

```{r}
install.packages("Hmisc")
library(Hmisc)
describe(tT$adj.P.Val)
nrow(tT)
```

```{r}
# UMAP plot (dimensionality reduction)
ex <- na.omit(ex) # eliminate rows with NAs
ex <- ex[!duplicated(ex), ]  # remove duplicates
ump <- umap(t(ex), n_neighbors = 15, random_state = 123)
par(mar=c(3,3,2,6), xpd=TRUE)
plot(ump$layout, main="UMAP plot, nbrs=15", xlab="", ylab="", col=gs, pch=20, cex=1.5)
legend("topright", inset=c(-0.15,0), legend=levels(gs), pch=20,
col=1:nlevels(gs), title="Group", pt.cex=1.5)
library("maptools")  # point labels without overlaps
pointLabel(ump$layout, labels = rownames(ump$layout), method="SANN", cex=0.6)

# mean-variance trend, helps to see if precision weights are needed
plotSA(fit2, main="Mean variance trend, GSE35977")

```



# Visualize and quality control test results.
```{r}

# Build histogram of P-values for all genes. Normal test
# assumption is that most genes are not differentially expressed.
tT2 <- topTable(fit2, adjust="fdr", sort.by="B", number=Inf)
hist(tT2$adj.P.Val, col = "grey", border = "white", xlab = "P-adj",
  ylab = "Number of genes", main = "P-adj value distribution")

```


DIFFERENTIALLY EXPRESSED GENES

```{r}
library(dplyr)
Degs_1 = filter(tT2, adj.P.Val<0.5)
nrow(Degs_1)
#head(Degs)



```

UPREGULATED GENES
```{r}
Up_Degs= filter(tT2, adj.P.Val<0.5 & (logFC)>0)
nrow(Up_Degs)

```
DOWN REGULATED GENES
```{r}
Down_Degs= filter(tT2, adj.P.Val<0.5 & (logFC)<0)
nrow(Down_Degs)
#Down_Degs$Gene.symbol
```
Saving into a diff variablr 
```{r}
degs_micro <- Degs_1$Gene.symbol
head(degs_micro)
```
```{r}

```

Top 10 DEGs and their KEGG PATHWAYS

```{r}
Deg_Sym <- Degs_1$Gene.symbol
length(Deg_Sym)

write.table(as.data.frame(Deg_Sym), file = "DEG_Micro.txt", col.names = FALSE,row.names = FALSE, quote = FALSE)
```
HEAT MAP
```{r}
library(gplots)
#heatmap.2(fit2, 
          #col = colorRampPalette(c("blue", "white", "red"))(100), 
          #scale = "row", 
          #dendrogram = "both", 
          #margins = c(10, 10))
mat <- as.matrix(ex)

# Remove missing values
mat <- na.omit(mat)

# Generate the heatmap
heatmap.2(mat, col = colorRampPalette(c("blue", "white", "red"))(100))
```

